A Data-Driven Framework for Neural Field Modeling

### 2011-01-15 ltx/first_order_framework/response_to_reviews.tex

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{soul}
\usepackage{anysize} %<-for setting margins
\marginsize {1.6cm}{1.6cm}{2.0cm}{2cm}
\usepackage{color}
\newcommand{\dean}[1]{\textcolor{red}{#1}}
\newcommand{\mike}[1]{\textcolor{green}{#1}}
\newcommand{\parham}[1]{\textcolor{blue}{#1}}

\title{Responses - A Data-Driven Framework for Neural Field Modeling}
\author{D. Freestone, P. Aram, M. Dewar, K. Scerri, D. Grayden, V. Kadirkamanathan}

@@

\begin{document}
    \maketitle

    The authors would very much like to thank the reviewers for their encouraging comments and their efforts to help improve this paper. Please find below the authors responses to the reviewers questions and suggestions regarding the manuscript. 
\\

Best Regards,
\\

Dean Freestone.

@@

    \end{enumerate}

    \subsection{Additional Comments}

    \begin{enumerate}
        \item The discrepancy of the results in Experiment II where the weights of the basis functions are not in agreement with the actual parameters while the reconstructed kernels seem satisfying.

		\emph{We have added to the results section describing experiment II: `These results indicate that the solution to the system may not be unique in non-ideal, practical situations. Nevertheless, the system is identifiable in terms of the net shape of the estimated kernel.'}

@@

        \item The use of Euler discretisation and how this affects the system's behavior.

		\emph{Amendments have been made to Appendix A, where the discretization is performed. We have provided a proof for the stability of the discrete-time system. We have also added a reference for the further analysis of the error introduced by Euler discretization method.}

@@

\emph{The cross-referencing has been corrected.}

    \end{enumerate}

    \section{Reviewer 2}

\begin{enumerate}
    \item The handling of the (to borrow a term from numerical weather forecasting) `sub-grid scale physics' is inadequate for nervous systems (see, e.g., equation 8 (now equation 12) [The nonlinear IntegroDifference Equation]). I do very much appreciate the spectral estimation of spatial and temporal scales. But the model errors are lumped into an iid model error term that is not realistic. For instance, the flow of potassium in the extracellular spaces has a known dynamic that will affect network excitability, generating important sub-grid dynamics that is anything but iid. See, for instance, the example in Ullah et al PLoS Comp Biol 2010 where without such dynamics, state tracking and parameter estimation for a neural system fails. A beautiful overview of sub-grid effects in data assimilation is found in Kalnay's 2003 book. I would very much value the authors comments on these issues. I guess were I to respond to this question, that I might consider extending the model error term to an autoregressive model. But I will let the authors answer this question.

	\emph{The authors believe that allowing for time-varying parameters will account for changes in excitability levels. In future work, we will also estimate the disturbance properties (spatial covariance and temporal variance). To address this point in this paper, we have added an extra paragraph to `Neural Field Model' section of the Methods, where we discuss the model mismatch and some basic modeling assumptions, such as stationarity of the parameters. In addition, we have added to the discussion (Extensions section) with possible ways to relax these assumptions.}

@@

\item Although we all know the 1970s version of Wilson and Cowan's field equations, there was an intriguing stochasticization of these equations by Benayoun et al PloS Comp Biol 2010. I and others will value the authors' thoughts of how a more palatable stochasticization of these field equations might influence the future development of the data-assimilation framework presented in this paper.

\emph{The results of Benayoun et al. provide evidence that the variance in a stochastic WC model will depend on the number of neurons that are modeled (in a stochastic network), where a larger network behaves more like the deterministic WC equations. This work encourages the use of the WC model for data-assimilation, but provides evidence that the disturbance properties should also be estimated. The authors believe that estimating the noise properties is possible, and we are planning to demonstrate how to achieve this in a subsequent paper. We have added some comments to the Extension to the Framework section in this regard.}

@@

\item What is the relation of the substitution of the temporal and spatial basis functions as employed in this paper to Galerkin's projection? How do the two techniques differ? If this is close, the non-local case for a Galerkin projection (typically used in diffusive processes) has not, to my knowledge, been done. Certainly not in a way that is obvious for neural field theory. Perhaps this is my ignorance. This seems important to clarify for the rest of us.

\emph{We have added an extra step in the derivation of the state-space model to highlight the relationship to our basis function decomposition and Galerkin's projection. We have also added an extra Appendix to further explain the relationship.}

\item In the isotropic and translationally invariant case examined here, would a Fourier basis for temporal and spatial structure have been a formally correct alternative?

\emph{Yes. Fourier basis functions would be a correct alternative to the case examined in this paper. In fact, a Fourier basis functions or any other orthonormal basis functions would simplify the state-space representation of the model, since $\Gamma$ (defined in equation (20)) of the manuscript would be the identity matrix. However, we have chosen Gaussian basis functions due to their semi-compact support. A comment has been added for possible alternative basis functions under equation (23) of the manuscript.}

\item In equation 29 (now equation (35)), what would a reasonable $\boldsymbol{\nu}_c$ be for a real neuronal system (mammalian) given the number of neurons and scale of electrodes? I am concerned that it might be much higher than any of us wish to consider. One would perhaps be saved by the filter characteristics of brain (e.g. Phys. Rev. E 73, 051911 (2006)).

@@

\emph{The mean spatial frequency content is actually lower than one might first expect. We have added `From the work of (Freeman et al, 2000), the expected bandwidth of the human neocortex is approximately 0.1-0.4 cycles/mm which requires a sensor spacing of approximately 1.25 mm to avoid distortions due to aliasing' to the Frequency Analysis section of the paper to clear this up. We have also added to the discussion. Note, we expect the spatial bandwidth of the cortex to be governed by the connectivity kernel, since it is equivalent to a spatial low-pass filter acting on the firing rates.}

\item The two covariance inflation parameters in equation 41 (now equation (47)) make me cringe. How are we best to adjust these fudge factors?

\emph{Values for the two parameters in equation 41 (now equation (47)) along with the relevant references, that describe the rationale for parameter choice, have been added to the manuscript.}

\item I understand that the Mexican Hat bases are in this paper often a composite of $\theta_0$, $\theta_1$, and $\theta_2$, representing short range excitation, short range inhibition, and long range excitation. But many will be confused by this subtlety, and at times in the text the Mexican Hat basis is described as if it were a single function. I would show how the theta's sum to create the general spatial basis shape of the hat in figure 1, and go through the paper in order to clarify this issue.

\emph{This point is clarified by adding the connectivity kernel components superimposed by the resulting Mexican Hat basis in Figure 1. of the manuscript. It has also been clarified throughout the paper.}
 % 
 % \begin{figure*}[!ht]
 % \begin{center}
 % \includegraphics{./Graph/pdf/fig1.pdf} 
 % \end{center}
 % \caption{}
 % \label{fig:Figure1}
 % \end{figure*}

### 2011-01-15 ltx/first_order_framework/LaTeX_Manuscript.tex
removed colors

% \newcommand{\dean}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}}
\newcommand{\dean}[1]{\textcolor{red}{#1}}
\newcommand{\parham}[1]{\textcolor{blue}{#1}}
\newcommand{\mike}[1]{\textcolor{Plum}{#1}}

@@

For models to be clinically viable, they must be patient-specific. A possible approach to achieve this would be to fit a general continuum neural field model, like the \citet{Wilson1973} (WC) or \citet{Amari1977} models, or a neural mass model like the \citet{Jansen1995} model, to patient-specific data. Fitting the neural models to individuals is a highly non-trivial task. Recently, however, this task has been approached from a number of standpoints. \dean{(Reviewer 1, additional comment d):) The first paper on patient-specific modeling (to the authors' best knowledge) came from~\citet{Valdes1999}, where they fit the neural mass model of \citet{LopesDaSilva1976} and \citet{Zetterberg1978} to EEG data using the local linearization filter \citep{Ozaki1993,Ozaki1994,Ozaki2000}. This paper demonstrated for the first time the feasibility of assimilating data with neural models of EEG.}

@@

\dean{(Reviewer 1, additional comment d):) Data-driven neural modeling was extended to the continuum approximation by \citet{Galka2008}, where they proposed an estimation framework based on a linear damped wave equation. Using a Kalman filter maximum likelihood framework, they improved on the low resolution electromagnetic tomography (LORETA) method for solving the inverse EEG problem. Another continuum neural field model-based estimator was developed by \citet{Daunizeau2009}, as an extension to the DCM framework. \parham{They argue that using a spatially extended continuum field model is superior for explaining cortical function than the neural mass DCM, since field models can explain a richer repertoire of dynamics such as traveling waves and bump solutions.}}

\dean{Another recent approach estimates the parameters of a modified WC neural field model using an unscented Kalman filter~\citep{schiff2008kalman}. This work takes a systems theoretic approach to the neural estimation problem, successfully demonstrating that it is possible to perform state estimation and control on spatiotemporal neural fields. This marks the first step in what has the potential to revolutionize the treatment of many neurological diseases where therapeutic electrical stimulation is viable. For other valuable contributions to data-driven neural modeling, see \citet{Nunez2000}, \citet{Jirsa2002}, and \citet{Robinson2004}.}

@@

In this paper, we demonstrate for the first time how an intracortical connectivity kernel can be inferred from data, based on a variant of the \citet{Wilson1973} neural field model. This work provides a fundamental link between the theoretical advances in neural field modeling and \dean{(for reviewer 1, comment 3:) high resolution intracranial} electrophysiological data. \dean{(for reviewer 1 comment 1:) (this is copied from the abstract) To illustrate the estimation framework, data is generated using the neural field equations incorporating modeled sensors enabling a comparison to be made between the estimated and true parameters.} The paper proceeds by first describing the continuum neural field equations that are used as the cortical model. Then a finite-dimensional neural field model is derived. The model is reduced by approximating the neural field using a set of continuous basis functions, weighted by a finite dimensional state vector. The next section establishes conditions, using spatial frequency analysis, for both sensor and basis function spacing and width, such that the dominant dynamics of the neural field can be represented by the reduced model. The state and parameter estimation procedure is described in the following section. The results for the spatial frequency analysis and parameter estimation are then presented. Finally, the implications and limitations of this framework are discussed along with planned future developments.

@@

\dean{(comment 1 and 9 from reviewer 2:) It is an implicit assumption that the neural field model (in equation~\ref{DiscreteTimeModel}) provides an apt description of the cortical dynamics recorded from a specific subject. Although models of this form are capable of describing a variety of cortical dynamics, there will be without doubt a mismatch between the cortex and the model. Nevertheless, there is a sufficient volume of interesting results from theoretical studies using the WC field equations that warrants the data assimilation framework. An advantage in using a lumped-parameter field model over a more detailed mathematical description is that the myriad of parameters that might influence excitability, such as specific ion concentrations, are considered to be lumped into parameters (connectivity kernel coefficients) that effectively describe the net system gain. Therefore, there are less parameters to estimate. A major challenge in model-based data analysis for the mass action of the brain is to make models sufficiently detailed, such that the parameters are meaningful, but simple enough to yield clear insights that can be related to theoretical studies (from both neuroscience and engineering perspectives).}

@@

The mapping between the membrane voltage and the electrophysiological data, denoted by $\mathbf{y}_t$, is modeled using the observation function that incorporates sensors with a spatial extent by
\begin{equation}\label{eq:ObservationEquation}
	y_t(\mathbf{r}_n) = \int_{\Omega} { m\left(\mathbf{r}_n-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, d\mathbf{r}'} + \varepsilon_t(\mathbf{r}_n), 
\end{equation}
where $m\left(\mathbf{r}_n-\mathbf{r}'\right)$ is the observation kernel, $\mathbf{r}_n$ defines the location of the electrodes in the field, $n=0,...,n_y-1$ indexes the sensors and $\varepsilon_t(\mathbf{r}_n) \sim \mathcal{N}\left(0,\boldsymbol{\Sigma}_{\varepsilon}\right)$ denotes a multivariate normal distribution with mean zero and the covariance matrix $\boldsymbol{\Sigma}_{\varepsilon} = \sigma_{\varepsilon}^2\mathbf{I}$. \dean{(for reviewer 1, comment 3:) Since we are considering intracranial measurements recorded directly from the surface of the cortex or within the brain, the lead field is not modeled by the observation equation.}

@@

The connectivity kernel can also be decomposed as 
\begin{equation}\label{DefKernelDecomp}
	 w\left(\mathbf{r},\mathbf{r}'\right) =\boldsymbol{\psi}^\top\left(\mathbf{r},\mathbf{r}'\right) \boldsymbol{\theta},
\end{equation}
where $\boldsymbol{\psi}(\mathbf{r},\mathbf{r}')$ is a vector of Gaussian basis functions and $\boldsymbol{\theta}$ is a vector of scaling parameters. \dean{A graphical demonstration of the connectivity kernel decomposition is provided in Fig.~\ref{fig:Figure1}.} By exploiting the isotropy of the connectivity basis functions, they can be written as  
$\boldsymbol{\psi}(\mathbf{r}-\mathbf{r}')$. We will assume that we know the parametric form of the connectivity basis functions, where the scaling parameters $\boldsymbol{\theta}$ are unknown. 
\dean{(for reviewer 1 point 2:) Note, that although the examples presented in this paper use the Mexican hat connectivity kernel, the framework is also valid for other connectivity kernels, provided they are homogeneous and that they can be approximated by a set of isotropic basis functions parametrized by a coefficient vector $\mathbf{\theta}$. The mathematical formulation does not restrict the shape of the kernel, where anisotropic arbitrary shapes can be represented. Furthermore, the kernel basis functions need not be a Gaussian shape, but can be, for example, Laplacian to describe the ``wizard hat'' connectivity or B-spline functions to describe a kernel with a compact support.}

Each connectivity basis function of the Mexican hat can, individually, be considered a layer in the Wilson and Cowan model. Rearranging the order of the spatial convolution in equation~\ref{DiscreteTimeModel} and substituting in equations~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} \dean{to the right hand side} we get the approximation of the dynamic field
% \begin{align}
% 	\label{reduced continuous model}
% 	\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_{t+1} &= T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}') \, d\mathbf{r}'}\boldsymbol{\theta} \nonumber \\
% 	&+ \xi\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_t + \mathbf{e}_t(\mathbf{r}). 
% \end{align}
\begin{align}
	\label{reduced continuous model}
	v_{t+1}\left(\mathbf{r}\right) &\approx T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}') \, d\mathbf{r}'}\boldsymbol{\theta} \nonumber \\
	&+ \xi\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_t + \mathbf{e}_t(\mathbf{r}). 
\end{align}
Next we cross-multiply equation~\ref{reduced continuous model} by $\boldsymbol{\phi}(\mathbf{r})$ and integrate over the spatial domain, $\Omega$, to get 
\begin{align}
	\label{eq:decompRHS}
 	\int_\Omega \boldsymbol{\phi}&\left(\mathbf{r}\right) v_{t+1}\left(\mathbf{r}\right)d\mathbf{r}  \nonumber \\
	&\approx T_s \int_\Omega \boldsymbol{\phi} (\mathbf{r}) \int_\Omega \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}') \mathbf{x}_t ) \, d\mathbf{r}'d\mathbf{r}\boldsymbol{\theta} \nonumber \\ 
	&+ \xi\int_\Omega {\boldsymbol{\phi}(\mathbf{r}) \boldsymbol{\phi}^{\top}(\mathbf{r})d\mathbf{r} } \mathbf{x}_t + \int_\Omega{\boldsymbol{\phi} (\mathbf{r}) e_t(\mathbf{r}) \, d\mathbf{r}}. 
\end{align}
\dean{The solution of this equation for $\mathbf{x}_t$ is equivalent to finding a Galerkin projection, where the field basis functions would be considered the projection functions and the states would be the amplitude coefficients. See~\ref{App:Galerkin} for more information.}

In the next step of forming the state-space model we substitute the field decomposition into the left hand side of equation~\ref{eq:decompRHS} giving
\begin{align}
	\label{eq:StartofReduction}
 	\int_\Omega \boldsymbol{\phi}&\left(\mathbf{r}\right) \boldsymbol{\phi}^{\top}\left(\mathbf{r}\right)\mathbf{x}_{t+1}d\mathbf{r}  \nonumber \\
	&= T_s \int_\Omega \boldsymbol{\phi} (\mathbf{r}) \int_\Omega \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}') \mathbf{x}_t ) \, d\mathbf{r}'d\mathbf{r}\boldsymbol{\theta} \nonumber \\ 
	&+ \xi\int_\Omega {\boldsymbol{\phi}(\mathbf{r}) \boldsymbol{\phi}^{\top}(\mathbf{r})d\mathbf{r} } \mathbf{x}_t + \int_\Omega{\boldsymbol{\phi} (\mathbf{r}) e_t(\mathbf{r}) \, d\mathbf{r}}. 
\end{align}
Now defining the matrix
\begin{equation}\label{eq:DefGamma}
	\boldsymbol{\Gamma} \triangleq \int_\Omega {\boldsymbol{\phi} \left(\mathbf{r}\right)\boldsymbol{\phi} ^{\top}\left(\mathbf{r}\right) \, d\mathbf{r}}, 
\end{equation}
and substituting it into equation~\ref{eq:StartofReduction} and cross-multiplying by $\boldsymbol{\Gamma}^{-1}$ gives
\begin{align}
    \label{eq:ReducedForm}
	 \mathbf{x}_{t+1} &= T_s\boldsymbol{\Gamma}^{-1}
	 \int_\Omega \boldsymbol{\phi}(\mathbf{r}) 
	 \int_\Omega \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}')f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t)d\mathbf{r}' d\mathbf{r} \boldsymbol{\theta}  \nonumber \\
	& + \xi\mathbf{x}_t + \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r}) e_t(\mathbf{r}) \, d\mathbf{r}}.
\end{align}
The analytic derivation of the inner product of \emph{n}-dimensional Gaussians, which was used for calculating $\mathbf\Gamma$, is provided in \ref{App:InnerProdOfGaussians}. \dean{Alternative field basis functions may be used to form the decomposition; if the basis is orthonormal, such as a Fourier basis, then $\Gamma$ is simply the identity matrix.} Equation~\ref{eq:ReducedForm} can be simplified by exploiting the isotropy of the connectivity kernel basis functions where
\begin{equation}
	\psi_i (\mathbf{r}-\mathbf{r}') = \psi_i (2\mathbf{c}_i+\mathbf{r}'-\mathbf{r})
\end{equation}
and $\mathbf{c}_i$ is the center of the $i^{th}$ basis function of the kernel decomposition. To make the simplification we define
\begin{equation}\label{eq:DefPsi}
	\left[ \boldsymbol\Psi(\mathbf{r}')\right]_{:i}  \triangleq T_s\boldsymbol{\Gamma}^{-1}\int_\Omega {\boldsymbol{\phi}(\mathbf{r})\psi_i (2\mathbf{c}_i+\mathbf{r}'-\mathbf{r})\textrm{d}\mathbf{r}},
\end{equation}
where $\left[ \boldsymbol\Psi(\mathbf{r}')\right]_{:i}$ denotes the $i^{th}$ column of $\boldsymbol{\Psi}(\mathbf{r}')$ which is a $n_x \times n_{\theta}$ matrix, where $n_x$ is the number of basis functions (states) and $n_{\theta}$ is the number of connectivity kernel basis functions. This matrix is pre-defined analytically (see \ref{ap:ConvOfGaussians} for analytic convolution of two Gaussians), since it is constant with respect to the states and the unknown parameters. Substituting $\boldsymbol{\Psi}(\mathbf{r}')$ into equation~\ref{eq:ReducedForm} gives
\begin{align}
	\mathbf{x}_{t+1} &= \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) \, d\mathbf{r}' \boldsymbol{\theta} \nonumber \\ &+ \xi\mathbf{x}_t 
+ \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r}) \, d\mathbf{r}}.
\end{align}
This substitution swaps the convolution in equation~\ref{reduced continuous model}, which is computationally demanding, with the convolution in equation~\ref{eq:DefPsi}, which is pre-defined analytically. This is of great importance as it lowers the computational demands of the estimation algorithm and provides a dramatic increase in estimation speed.

>>>>

% \newcommand{\dean}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}}
% \newcommand{\dean}[1]{\textcolor{red}{#1}}
% \newcommand{\parham}[1]{\textcolor{blue}{#1}}
% \newcommand{\mike}[1]{\textcolor{Plum}{#1}}

@@

For models to be clinically viable, they must be patient-specific. A possible approach to achieve this would be to fit a general continuum neural field model, like the \citet{Wilson1973} (WC) or \citet{Amari1977} models, or a neural mass model like the \citet{Jansen1995} model, to patient-specific data. Fitting the neural models to individuals is a highly non-trivial task. Recently, however, this task has been approached from a number of standpoints. The first paper on patient-specific modeling (to the authors' best knowledge) came from~\citet{Valdes1999}, where they fit the neural mass model of \citet{LopesDaSilva1976} and \citet{Zetterberg1978} to EEG data using the local linearization filter \citep{Ozaki1993,Ozaki1994,Ozaki2000}. This paper demonstrated for the first time the feasibility of assimilating data with neural models of EEG.

@@

Data-driven neural modeling was extended to the continuum approximation by \citet{Galka2008}, where they proposed an estimation framework based on a linear damped wave equation. Using a Kalman filter maximum likelihood framework, they improved on the low resolution electromagnetic tomography (LORETA) method for solving the inverse EEG problem. Another continuum neural field model-based estimator was developed by \citet{Daunizeau2009}, as an extension to the DCM framework. They argue that using a spatially extended continuum field model is superior for explaining cortical function than the neural mass DCM, since field models can explain a richer repertoire of dynamics such as traveling waves and bump solutions.

Another recent approach estimates the parameters of a modified WC neural field model using an unscented Kalman filter~\citep{schiff2008kalman}. This work takes a systems theoretic approach to the neural estimation problem, successfully demonstrating that it is possible to perform state estimation and control on spatiotemporal neural fields. This marks the first step in what has the potential to revolutionize the treatment of many neurological diseases where therapeutic electrical stimulation is viable. For other valuable contributions to data-driven neural modeling, see \citet{Nunez2000}, \citet{Jirsa2002}, and \citet{Robinson2004}.

@@

In this paper, we demonstrate for the first time how an intracortical connectivity kernel can be inferred from data, based on a variant of the \citet{Wilson1973} neural field model. This work provides a fundamental link between the theoretical advances in neural field modeling and high resolution intracranial electrophysiological data. To illustrate the estimation framework, data is generated using the neural field equations incorporating modeled sensors enabling a comparison to be made between the estimated and true parameters. The paper proceeds by first describing the continuum neural field equations that are used as the cortical model. Then a finite-dimensional neural field model is derived. The model is reduced by approximating the neural field using a set of continuous basis functions, weighted by a finite dimensional state vector. The next section establishes conditions, using spatial frequency analysis, for both sensor and basis function spacing and width, such that the dominant dynamics of the neural field can be represented by the reduced model. The state and parameter estimation procedure is described in the following section. The results for the spatial frequency analysis and parameter estimation are then presented. Finally, the implications and limitations of this framework are discussed along with planned future developments.

@@

It is an implicit assumption that the neural field model (in equation~\ref{DiscreteTimeModel}) provides an apt description of the cortical dynamics recorded from a specific subject. Although models of this form are capable of describing a variety of cortical dynamics, there will be without doubt a mismatch between the cortex and the model. Nevertheless, there is a sufficient volume of interesting results from theoretical studies using the WC field equations that warrants the data assimilation framework. An advantage in using a lumped-parameter field model over a more detailed mathematical description is that the myriad of parameters that might influence excitability, such as specific ion concentrations, are considered to be lumped into parameters (connectivity kernel coefficients) that effectively describe the net system gain. Therefore, there are less parameters to estimate. A major challenge in model-based data analysis for the mass action of the brain is to make models sufficiently detailed, such that the parameters are meaningful, but simple enough to yield clear insights that can be related to theoretical studies (from both neuroscience and engineering perspectives).

@@

The mapping between the membrane voltage and the electrophysiological data, denoted by $\mathbf{y}_t$, is modeled using the observation function that incorporates sensors with a spatial extent by
\begin{equation}\label{eq:ObservationEquation}
	y_t(\mathbf{r}_n) = \int_{\Omega} { m\left(\mathbf{r}_n-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, d\mathbf{r}'} + \varepsilon_t(\mathbf{r}_n), 
\end{equation}
where $m\left(\mathbf{r}_n-\mathbf{r}'\right)$ is the observation kernel, $\mathbf{r}_n$ defines the location of the electrodes in the field, $n=0,...,n_y-1$ indexes the sensors and $\varepsilon_t(\mathbf{r}_n) \sim \mathcal{N}\left(0,\boldsymbol{\Sigma}_{\varepsilon}\right)$ denotes a multivariate normal distribution with mean zero and the covariance matrix $\boldsymbol{\Sigma}_{\varepsilon} = \sigma_{\varepsilon}^2\mathbf{I}$. Since we are considering intracranial measurements recorded directly from the surface of the cortex or within the brain, the lead field is not modeled by the observation equation.

@@

The connectivity kernel can also be decomposed as 
\begin{equation}\label{DefKernelDecomp}
	 w\left(\mathbf{r},\mathbf{r}'\right) =\boldsymbol{\psi}^\top\left(\mathbf{r},\mathbf{r}'\right) \boldsymbol{\theta},
\end{equation}
where $\boldsymbol{\psi}(\mathbf{r},\mathbf{r}')$ is a vector of Gaussian basis functions and $\boldsymbol{\theta}$ is a vector of scaling parameters. A graphical demonstration of the connectivity kernel decomposition is provided in Fig.~\ref{fig:Figure1}. By exploiting the isotropy of the connectivity basis functions, they can be written as  
$\boldsymbol{\psi}(\mathbf{r}-\mathbf{r}')$. We will assume that we know the parametric form of the connectivity basis functions, where the scaling parameters $\boldsymbol{\theta}$ are unknown. 
Note, that although the examples presented in this paper use the Mexican hat connectivity kernel, the framework is also valid for other connectivity kernels, provided they are homogeneous and that they can be approximated by a set of isotropic basis functions parametrized by a coefficient vector $\mathbf{\theta}$. The mathematical formulation does not restrict the shape of the kernel, where anisotropic arbitrary shapes can be represented. Furthermore, the kernel basis functions need not be a Gaussian shape, but can be, for example, Laplacian to describe the ``wizard hat'' connectivity or B-spline functions to describe a kernel with a compact support.

Each connectivity basis function of the Mexican hat can, individually, be considered a layer in the Wilson and Cowan model. Rearranging the order of the spatial convolution in equation~\ref{DiscreteTimeModel} and substituting in equations~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} to the right hand side we get the approximation of the dynamic field
\begin{align}
	\label{reduced continuous model}
	v_{t+1}\left(\mathbf{r}\right) &\approx T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}') \, d\mathbf{r}'}\boldsymbol{\theta} \nonumber \\
	&+ \xi\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_t + \mathbf{e}_t(\mathbf{r}). 
\end{align}
Next we cross-multiply equation~\ref{reduced continuous model} by $\boldsymbol{\phi}(\mathbf{r})$ and integrate over the spatial domain, $\Omega$, to get 
\begin{align}
	\label{eq:decompRHS}
 	\int_\Omega \boldsymbol{\phi}&\left(\mathbf{r}\right) v_{t+1}\left(\mathbf{r}\right)d\mathbf{r}  \nonumber \\
	&\approx T_s \int_\Omega \boldsymbol{\phi} (\mathbf{r}) \int_\Omega \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}') \mathbf{x}_t ) \, d\mathbf{r}'d\mathbf{r}\boldsymbol{\theta} \nonumber \\ 
	&+ \xi\int_\Omega {\boldsymbol{\phi}(\mathbf{r}) \boldsymbol{\phi}^{\top}(\mathbf{r})d\mathbf{r} } \mathbf{x}_t + \int_\Omega{\boldsymbol{\phi} (\mathbf{r}) e_t(\mathbf{r}) \, d\mathbf{r}}. 
\end{align}
The solution of this equation for $\mathbf{x}_t$ is equivalent to finding a Galerkin projection, where the field basis functions would be considered the projection functions and the states would be the amplitude coefficients. See~\ref{App:Galerkin} for more information. In the next step of forming the state-space model we substitute the field decomposition into the left hand side of equation~\ref{eq:decompRHS} giving
\begin{align}
	\label{eq:StartofReduction}
 	\int_\Omega \boldsymbol{\phi}&\left(\mathbf{r}\right) \boldsymbol{\phi}^{\top}\left(\mathbf{r}\right)\mathbf{x}_{t+1}d\mathbf{r}  \nonumber \\
	&= T_s \int_\Omega \boldsymbol{\phi} (\mathbf{r}) \int_\Omega \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}') \mathbf{x}_t ) \, d\mathbf{r}'d\mathbf{r}\boldsymbol{\theta} \nonumber \\ 
	&+ \xi\int_\Omega {\boldsymbol{\phi}(\mathbf{r}) \boldsymbol{\phi}^{\top}(\mathbf{r})d\mathbf{r} } \mathbf{x}_t + \int_\Omega{\boldsymbol{\phi} (\mathbf{r}) e_t(\mathbf{r}) \, d\mathbf{r}}. 
\end{align}
Now defining the matrix
\begin{equation}\label{eq:DefGamma}
	\boldsymbol{\Gamma} \triangleq \int_\Omega {\boldsymbol{\phi} \left(\mathbf{r}\right)\boldsymbol{\phi} ^{\top}\left(\mathbf{r}\right) \, d\mathbf{r}}, 
\end{equation}
and substituting it into equation~\ref{eq:StartofReduction} and cross-multiplying by $\boldsymbol{\Gamma}^{-1}$ gives
\begin{align}
    \label{eq:ReducedForm}
	 \mathbf{x}_{t+1} &= T_s\boldsymbol{\Gamma}^{-1}
	 \int_\Omega \boldsymbol{\phi}(\mathbf{r}) 
	 \int_\Omega \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}')f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t)d\mathbf{r}' d\mathbf{r} \boldsymbol{\theta}  \nonumber \\
	& + \xi\mathbf{x}_t + \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r}) e_t(\mathbf{r}) \, d\mathbf{r}}.
\end{align}
The analytic derivation of the inner product of \emph{n}-dimensional Gaussians, which was used for calculating $\mathbf\Gamma$, is provided in \ref{App:InnerProdOfGaussians}. Alternative field basis functions may be used to form the decomposition; if the basis is orthonormal, such as a Fourier basis, then $\Gamma$ is simply the identity matrix. Equation~\ref{eq:ReducedForm} can be simplified by exploiting the isotropy of the connectivity kernel basis functions where
\begin{equation}
	\psi_i (\mathbf{r}-\mathbf{r}') = \psi_i (2\mathbf{c}_i+\mathbf{r}'-\mathbf{r})
\end{equation}
and $\mathbf{c}_i$ is the center of the $i^{th}$ basis function of the kernel decomposition. To make the simplification we define
\begin{equation}\label{eq:DefPsi}
	\left[ \boldsymbol\Psi(\mathbf{r}')\right]_{:i}  \triangleq T_s\boldsymbol{\Gamma}^{-1}\int_\Omega {\boldsymbol{\phi}(\mathbf{r})\psi_i (2\mathbf{c}_i+\mathbf{r}'-\mathbf{r})\textrm{d}\mathbf{r}},
\end{equation}
where $\left[ \boldsymbol\Psi(\mathbf{r}')\right]_{:i}$ denotes the $i^{th}$ column of $\boldsymbol{\Psi}(\mathbf{r}')$ which is a $n_x \times n_{\theta}$ matrix, where $n_x$ is the number of basis functions (states) and $n_{\theta}$ is the number of connectivity kernel basis functions. This matrix is pre-defined analytically (see \ref{ap:ConvOfGaussians} for analytic convolution of two Gaussians), since it is constant with respect to the states and the unknown parameters. Substituting $\boldsymbol{\Psi}(\mathbf{r}')$ into equation~\ref{eq:ReducedForm} gives
\begin{align}
	\mathbf{x}_{t+1} &= \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) \, d\mathbf{r}' \boldsymbol{\theta} \nonumber \\ &+ \xi\mathbf{x}_t 
+ \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r}) \, d\mathbf{r}}.
\end{align}
This substitution swaps the convolution in equation~\ref{reduced continuous model}, which is computationally demanding, with the convolution in equation~\ref{eq:DefPsi}, which is pre-defined analytically. This is of great importance as it lowers the computational demands of the estimation algorithm and provides a dramatic increase in estimation speed.
